---
title: "Dealing with Complex data"
subtitle: "Georgia Tech, humdrumR Workshop"
author: "Nat Condit-Schultz"
date: "May 11, 2023"
output:
  html_document:
    df_print: paged
    toc: true
    toc_float: true
    theme: flatly
editor_options: 
  chunk_output_type: console
---

```{r}
library(humdrumR)
```


The humdrum syntax is very flexible. There are lots of ways that data can be organized.

+ We can have multiple aspects of musical scrunched into single strings (`"[4.ee-/JJ"`]).
+ Or we can spread information across...
  + multiple spines
  + spine paths
  + multi-stops
+ We can also put information into tandem interpretations...

When dealing with humdrum data, we have to figure out how to make sense of any of these possibilities.
There is no general solution that humdrumR can automatically do for you!

----


When humdrumR reads in humdrum data, it reads it into a structurally agnostic "flat" table.
Each and every token gets its own row.
Sometimes, this is confusing, and/or not what you want!



# Working with subsets of data

Let's consider the CoCoPops Billboard dataset:

```{r}

bb <- readHumdrum('CoCoPops-Billboard/.*hum')

```

Take a look at this dataset---many of the files are too big to fit on our screen!
Let's use `spines()`:

```{r}
interpretations(bb)


```




Our files have one to five pairs of `**kern` and `**silbe` spines, followed by `**harmony`, `chordsym`, `**phrase`...etc.


Many (most) humdrumR functions automatically look at the `Exclusive` field to determine where/when/if they should be applied.
This means, that functions will be fairly smart about only working on appropriate data.
For example,

```{r}
within(bb, kern(Token))

```


The `kern()` function works on the `**kern` spines, but ignores the other spines. They all return `NA` (so we see null tokens).

-----


This is nice, but is still generally a good practice to explicitly remove/avoid data we are not interested in.
One way is to provide `with/within` a subset argument.

```{r}

within(bb, subset = Exclusive == 'kern', kern(Token))

```

In this case, we get the exactly the same result as before, we are just being more explicit about things.
This is going to be a safer approach in general!

Notice that when you give subset as an *argument* to `within`, the command you run is applied to the subset, but the rest of he data is not removed from the object.



## Discarding data

Now, in many cases, we might have information in our data that we simply don't care about at all---depending on our research question.
In this case, the best practice is just to get rid of it!
For example we can extract only the spines we are interested in, using `Exclusive` field perhaps, and overwrite our `bb`.
This should generally be the first thing you do after reading data in: get rid of anything you don't want.

```{r}
subset(bb, Exclusive %in% c('kern', 'silbe', 'harmony')) |> removeEmptySpines() -> bb

bb

```


Ok, that's much more manegable!

# Folding data

Ok, we have our kern tokens (which include rhythm and pitch data), lyrics, and harmony.
But remember, at this point each and every token is it's own row in the humdrum table.
So the `Token` field includes both kern, silbe, and harmony, all together!

```{r}
getHumtab(bb, 'D')[ , c('Token', 'File', 'Spine', 'Record', 'Key', 'Bar', 'Exclusive')][Bar == 8][1:40]

```


Let's review the idea of 'fields' for a second.
You can think of the fields in a humdrumR object as a third dimension in the humdrum data. 
"Layers" that are overlaid on top of each other.
You can also think of them as different "viewpoints" of the same data.

```{r}
bb[1]$Key
bb[1]$Spine
bb[1]$COC

```

## Folding

We currently have three (types of) spines of data.
The data kern/harm/silbe data points are at different "locations" in the humdrum data---different rows in the underlying table.
We (might) want to "move" these data points "on top" of each other, into new fields/layers.
This makes the data align in a way that is mor useful.

We call this "folding" the humdrum data.
Let's try it with the lyrics:

```{r}
bb <- foldExclusive(bb, 'silbe', 'kern')

```

The `**silbe` spines have been removed from `Token`, and moved over into a new field (`Silbe`).
Notice that each `**silbe` spine is paired with the adjacent `**kern` spine.

We can next do the same thing with the `**harmony` spine:


```{r}
bb <- foldExclusive(bb$Token, 'harmony', 'kern')

```

If there are more `**kern` spines than `**harmony` spines, the harmonies are duplicated!


---

We now have our lyrics, harmony, and kern all "aligned" for ideal analysis.
We might want to split the pitch and rhythm informnation:

```{r}
bb <- within(bb$Token, Kern <- kern(Token), Recip <- recip(untie(Token)))


```




## Ditto

You probably noticed that the harmony spine doesn't actually have chords aligned with every note.
If we want to "fill forward" data into subsequent null data tokens, we can use ditto:

```{r}
bb <- within(bb$Token, Harmony <- ditto(tertian(Harmony)))
```



# Paths and Stops


HumdrumR parses spine paths, numbering them (within parent spines) from left to right, starting from 0, in the `Path` field.
HumdrumR parses all "multi-stops" (a.k.a., sub-tokens) as separate tokens, which are numbers in the `Stop` field, starting from 1.


```{r}
tavern <- readHumdrum('TAVERN/.*krn')


```


When analyzing music like this, its not at all clear how we *want* to treat spine paths and/or multistops.
The laziest/easiest solution, is just to remove them:


```{r}
subset(tavern, Path == 0 & Stop == 1) |> removeEmptyPaths() |> removeEmptyStops() -> tavern


```


## Expanding paths

We can "fill out" spine paths using the `expandPaths()` function.

```{r}
expandPaths(tavern)

```
