---
title: "humdrumRworkshop_practiceworksheet"
---
Load library
```{r}
library(humdrumR)
```

# Chorales

> Read-in the Chorales corpus, and save the corpus.

```{r}


```


### Rhythm

> Find the distribution (counts) of all rhythmic values in the corpus (hint: use the tally() function); optionally (bar)plot as well.
> How often does Bach use 16th notes? How often does he use half-notes? What other rhythmic values does he use?

```{r}

```

### Scale degrees

> Plot (barplot) the distribution of scale degrees in the corpus; then plot the distribution of scale degrees separately for major and minor-mode pieces.


```{r}


```

### Melodic Intervals

> Look at the distribution of melodic intervals in the Soprano part.

```{r}

```


> Which voice uses the most unisons? Which voice uses the most leaps?


```{r}


```

### Keys


> Which keys does Bach use the most (and the least)?

```{r}

```



# Chorales With Harmony

> Load the corpus from the "WithHarm" subdirectory and save.

```{r}


```

### Cadences

> What chords most commonly end a phrase (i.e., fermata) in the Choralebh corpus?
> Hint 1: you need to index to the first column, or, subset to only harm spine
> Hint 2: fermatas are represented by ";"

```{r}


```


> What are the top five most common roman numerals in major keys?
> Note: The chorale data only includes one global key indication, no indications of modulations.

```{r}


```



